Nondeterministic Bigraphical Reactive Systems for Markov Decision Processes\thanks{Supported by organization x.}

### 2018-08-13 paper.tex
TODOs + MDP references + a prettier door

\begin{abstract}
  In this paper we introduce two extensions of bigraphical reactive systems
  (BRSs) to handle probabilistic and nondeterministic behaviour, allowing BRSs
  to be used in discrete-time Markov chain and Markov decision process (MDP)
  generation. We extend the implementation of an open-source tool for BRSs
  BigraphER with new syntax for MDP actions and rewards, better visualisation
  capabilities, and data export options. Furthermore, we introduce a new
  front-end for working with BRSs based on Jupyter notebooks, allowing the user
  to quickly visualise bigraphs and reaction rules, simulate execution, or
  produce a complete transition system, all within the notebook interface.
  Finally, we demonstrate the modelling capabilities of the new system with a
  case study on the movement and behaviour of autonomous agents, exploring ideas
  such as collecting objects, tracking visited locations, hierarchies of spaces,
  and uncertainty in the topology of space.

@@

\begin{definition}[Markov decision process] \label{mdp}
  For any finite set $X$, let $\Dist(X)$ denote the set of discrete probability
  distributions over $X$. A \emph{Markov Decision Process} is a tuple $ (S,
  \overline{s}, A, P, L)$, where: $S$ is a finite set of states and
  $\overline{s} \in S$ is the initial state; $A$ is a finite set of
  \emph{actions}; $P : S \times A \to \Dist(S)$ is a (partial) probabilistic
  transition function, mapping state-action pairs to probability distributions
  over $S$; $L : S \to 2^P$ is a labelling with atomic propositions.
\end{definition}

@@

@@

\begin{figure}
  \centering
  \includegraphics{models/example_bigraph/foo_lhs.pdf}
  \caption{An example bigraph.}
  \label{example_bigraph}
\end{figure}

@@

\begin{definition}[concrete place graph with sharing]
  A \emph{concrete place graph with sharing}
  \[ F = (V_F, \ctrl_F, \prnt_F) : m \to n \]
  is a triple having an inner interface $m$ and an outer interface $n$. These
  index the sites and regions of the place graph, respectively. $F$ has a finite
  set $V_F \subset \mathcal{V}$ of nodes, a control map $\ctrl_F : V_F \to
  \mathcal{K}$, and a \emph{parent relation}
  \[ \prnt_F \subseteq (m \uplus V_F) \times (V_F \uplus n) \]
  that is acyclic, i.e., $(v, v) \not\in \prnt_F^+$ for any $v \in V_F$.
\end{definition}

@@

\begin{definition}[reaction rule] \label{reaction_rule}
  A \emph{reaction rule} is a pair
  \[ \mathsf{R} = (R : m \to J, R' : m \to J), \]
  sometimes written as $R \longrightarrowRHD R'$, where $R$ is the \emph{redex}
  and $R'$ the \emph{reactum}, and $R$ is solid. The rule generates all the
  \emph{ground reaction rules} $(r, r')$, where $r = (R \otimes \id_Y) \circ d$
  and $r' = (R' \otimes \id_Y) \circ d$ for some discrete ground
  \emph{parameter} $d : \epsilon \to \langle m, Y \rangle$. The \emph{reaction
    relation} $\longrightarrowrhd_{\mathsf{R}}$ over ground bigraphs is defined
  by
  \[ g \longrightarrowrhd_{\mathsf{R}} g' \text{ iff } g = D \circ r \text{ and
    } g' = D \circ r' \]
  for some bigraph $D$ and some ground reaction rule $(r, r')$ generated from
  $\mathsf{R}$.
\end{definition}

@@

\begin{definition}[probabilistic reaction rule]
  A \emph{probabilistic reaction rule} $\mathsf{R}$ is a triple $(R, R', p)$,
  sometimes written $R \xrightarrowRHD{p} R'$, where $(R, R')$ is a reaction
  rule and $p \in (0, 1]$ is a probability. Similarly to
  Definition~\ref{reaction_rule}, it generates a set of ground reaction rules of
  the form $(r, r', p)$.
\end{definition}

@@

@@

\begin{definition}[nondeterministic bigraphical reactive system (NBRS)]
  A \emph{nondeterministic bigraphical reactive system} consists of a pair
  $(\mathcal{B}, \mathcal{R})$, where $\mathcal{B}$ is a set of ground bigraphs
  and $\mathcal{R}$ is a set of nondeterministic reaction rules defined over
  $\mathcal{B}$.

@@

@@

@@

% TODO xavier \cite{koenig1998xavier}

@@

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    % walls
    \draw (3, 2) -- (3, 0) -- (1, 0) -- (1, 2) -- (0, 2) -- (0, 3) -- (3, 3) -- (3, 2) -- (2, 2);
    \draw[brown,ultra thick] (1, 2) -- (2, 2);
    % the door
    \draw[red,ultra thick,dashed] (2, 2) -- (2, 1);
    % the background grid
    \draw[gray,very thin] (1, 3) -- (1, 2);
    \draw[gray,very thin] (2, 3) -- (2, 2);
    \draw[gray,very thin] (1, 1) -- (3, 1);
    \draw[gray,very thin] (2, 1) -- (2, 0);
    % the agent
    \node at (0.5, 2.3) {\PHchild};
  \end{tikzpicture}
  \caption{The full map. The agent is in the top left cell. There is a thick
    brown door connecting the middle cell of the top row with a $2 \times 2$
    room below. The red dashed wall exists with probability $1-p$.}
  \label{agent2_map}
\end{figure}

@@

@@

% TODO advantages: more visual/realistic than a vector of numbers, rewards and
% transition rules can be generated with a higher degree of automation, less
% code, more flexible (changing size of the grid); disadvantages: constant
% probability (without hacking)

>>>>

% TODO: opening statement on why probabilities and non-determinism are important
\begin{abstract}
  In this paper we introduce two extensions of bigraphical reactive systems
  (BRSs) to handle probabilistic and nondeterministic behaviour, allowing BRSs
  to be used in discrete-time Markov chain and Markov decision process (MDP)
  generation. We extend the implementation of an open-source tool for BRSs
  BigraphER with new syntax for MDP actions and rewards, better visualisation
  capabilities, and data export options. Furthermore, we introduce a new
  front-end for working with BRSs based on Jupyter notebooks, allowing the user
  to quickly visualise bigraphs and reaction rules, simulate execution, or
  produce a complete transition system, all within the notebook interface.
  Finally, we demonstrate the modelling capabilities of the new system with a
  case study on the movement and behaviour of autonomous agents, exploring ideas
  such as collecting objects, tracking visited locations, hierarchies of spaces,
  and uncertainty in the topology of space.

@@

% TODO: short intro on MDPs with references to the literature, why they are
% important
% TODO: turn the door into an actual door

MDPs have been used to solve a wide range of problems: optimising airline yield
management \cite{DBLP:journals/transci/LautenbacherS99}, constructing optimal
dialogue strategies \cite{658989,DBLP:conf/icassp/LevinPE98}, electricity supply
bidding \cite{867150}, and many others.

% TODO xavier \cite{koenig1998xavier}
\begin{definition}[Markov decision process] \label{mdp}
  For any finite set $X$, let $\Dist(X)$ denote the set of discrete probability
  distributions over $X$. A \emph{Markov Decision Process} is a tuple $ (S,
  \overline{s}, A, P, L)$, where: $S$ is a finite set of states and
  $\overline{s} \in S$ is the initial state; $A$ is a finite set of
  \emph{actions}; $P : S \times A \to \Dist(S)$ is a (partial) probabilistic
  transition function, mapping state-action pairs to probability distributions
  over $S$; $L : S \to 2^P$ is a labelling with atomic propositions.
\end{definition}

@@

% TODO: minimal example with a diagram

@@

% TODO: site should be at the bottom
\begin{figure}
  \centering
  \includegraphics{models/example_bigraph/foo_lhs.pdf}
  \caption{An example bigraph.}
  \label{example_bigraph}
\end{figure}

@@

% TODO: mention we will be using bigraphs with sharing and we recall some
% definitions from TCS paper
\begin{definition}[concrete place graph with sharing]
  A \emph{concrete place graph with sharing}
  \[ F = (V_F, \ctrl_F, \prnt_F) : m \to n \]
  is a triple having an inner interface $m$ and an outer interface $n$. These
  index the sites and regions of the place graph, respectively. $F$ has a finite
  set $V_F \subset \mathcal{V}$ of nodes, a control map $\ctrl_F : V_F \to
  \mathcal{K}$, and a \emph{parent relation}
  \[ \prnt_F \subseteq (m \uplus V_F) \times (V_F \uplus n) \]
  that is acyclic, i.e., $(v, v) \not\in \prnt_F^+$ for any $v \in V_F$.
\end{definition}

@@

% TODO: why we limit reaction rules to solid (stochastic bigraphs paper)
\begin{definition}[reaction rule] \label{reaction_rule}
  A \emph{reaction rule} is a pair
  \[ \mathsf{R} = (R : m \to J, R' : m \to J), \]
  sometimes written as $R \longrightarrowRHD R'$, where $R$ is the \emph{redex}
  and $R'$ the \emph{reactum}, and $R$ is solid. The rule generates all the
  \emph{ground reaction rules} $(r, r')$, where $r = (R \otimes \id_Y) \circ d$
  and $r' = (R' \otimes \id_Y) \circ d$ for some discrete ground
  \emph{parameter} $d : \epsilon \to \langle m, Y \rangle$. The \emph{reaction
    relation} $\longrightarrowrhd_{\mathsf{R}}$ over ground bigraphs is defined
  by
  \[ g \longrightarrowrhd_{\mathsf{R}} g' \text{ iff } g = D \circ r \text{ and
    } g' = D \circ r' \]
  for some bigraph $D$ and some ground reaction rule $(r, r')$ generated from
  $\mathsf{R}$.
\end{definition}

@@

% TODO: for the same D_i - not clear
% TODO: minimal example showing normalisation
\begin{definition}[probabilistic reaction rule]
  A \emph{probabilistic reaction rule} $\mathsf{R}$ is a triple $(R, R', p)$,
  sometimes written $R \xrightarrowRHD{p} R'$, where $(R, R')$ is a reaction
  rule and $p \in (0, 1]$ is a probability. Similarly to
  Definition~\ref{reaction_rule}, it generates a set of ground reaction rules of
  the form $(r, r', p)$.
\end{definition}

@@

% TODO: intro explaining how we are further generalising the extensions defined
% above

@@

% TODO: D_i not clear
% TODO: different symbol for reward
\begin{definition}[nondeterministic bigraphical reactive system (NBRS)]
  A \emph{nondeterministic bigraphical reactive system} consists of a pair
  $(\mathcal{B}, \mathcal{R})$, where $\mathcal{B}$ is a set of ground bigraphs
  and $\mathcal{R}$ is a set of nondeterministic reaction rules defined over
  $\mathcal{B}$.

@@

% TODO: minimal example, possible the same one as below

@@

% TODO: we extend BigraphER
% TODO: describe the new syntactical constructs using the example
% TODO: what's the reward of action b?

@@

% TODO: the agent is simply modelled as an atomic node as the focus of this case
% study is on its movements across a grid. More complex models are possible
% by...

@@

% TODO: taking advantage of the hierarchical structure of bigraphs

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    % outside walls
    \draw (3, 2) -- (3, 0) -- (1, 0) -- (1, 2) -- (0, 2) -- (0, 3) -- (3, 3) -- (3, 2) -- (2, 2);
    % the probabilistic wall
    \draw[red,ultra thick,dashed] (2, 2) -- (2, 1);
    % the background grid
    \draw[gray,very thin] (1, 3) -- (1, 2);
    \draw[gray,very thin] (2, 3) -- (2, 2);
    \draw[gray,very thin] (1, 1) -- (3, 1);
    \draw[gray,very thin] (2, 1) -- (2, 0);
    % the agent
    \node at (0.5, 2.3) {\PHchild};
    % the door
    \draw[dotted] (1.25,2) arc (180:270:0.5cm);
    \draw[brown,ultra thick] (1.75,1.5) -- (1.75,2) node[right,  near start]{};
    \draw (1, 2) -- (1.25, 2);
    \draw (1.75, 2) -- (2, 2);
  \end{tikzpicture}
  \caption{The full map. The agent is in the top left cell. There is a thick
    brown door connecting the middle cell of the top row with a $2 \times 2$
    room below. The red dashed wall exists with probability $1-p$.}
  \label{agent2_map}
\end{figure}

@@

% TODO: limitation of a single entrance/exit per cell, can a room have multiple
% doors (doesn't make sense for this conceptualisation of the problem)?

% TODO: state diagrams (and statistics about them)
% TODO: numerical analysis (or ideas)
% TODO: discussion section
% TODO: what are the adtantages of using NBRS?
% TODO: limitations
% TODO: difficulties

@@

% TODO advantages: more visual/realistic than a vector of numbers, rewards and
% transition rules can be generated with a higher degree of automation, less
% code, more flexible (changing size of the grid); disadvantages: constant
% probability (without hacking)
% Extensions: multiple agents (nice), uncertainty in actions (easy), rotations
% (pomdp paper, easy), not limited to four directions
